---
title: "Mapping Metadata to Aesthetics"
---

```{r, include = FALSE}
library(rbiom)
pals <- sapply(simplify = FALSE, names(rbiom:::PALETTES), function (i) {
  sprintf("<td align='right' style='padding-left:20px; padding-right:0'>%s</td><td>%s</td>", i, 
    paste0(collapse = '', "<span style='color:#", rbiom:::PALETTES[[i]], "'>&#9608;</span>") )})

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



Metadata can be used to flexibly subset, partition, and apply aesthetics 
when creating a plot.


# Syntax examples


``` r
## Colors ----------------------------
color.by = "Body Site"
color.by = list('Body Site' = "bright")
color.by = list('Body Site' = c("Stool", "Saliva"))
color.by = list('Body Site' = list('values' = c("Stool", "Saliva"), 'colors' = "tol"))
color.by = list('Body Site' = c('Stool' = "blue", 'Saliva' = "green"))

## Patterns --------------------------
pattern.by = "Body Site"
pattern.by = list('Body Site' = c("Stool", "Saliva"))
pattern.by = list('Body Site' = c('Stool' = "left45", 'Saliva' = "hs_cross"))

## Shapes ----------------------------
shape.by = "Body Site"
shape.by = list('Body Site' = c("Stool", "Saliva"))
shape.by = list('Body Site' = c('Stool' = 7, 'Saliva' = 8))

## Facets ----------------------------
facet.by = "Body Site"
facet.by = c("Body Site", "Sex")
facet.by = list('Body Site' = c("Stool", "Saliva"), "Sex")

## Limits ----------------------------
limit.by = list('Sex' = "Male", 'Age' = c(20,40))
limit.by = list('Body Site' = c("Saliva", "Anterior nares"), 'Age' = c(NA,35))
```




# Colors


## Automatic colors

All `color.by` needs is the name of the metadata field you'd like to color by.
Specifying a subset can be done here too. In these cases, rbiom will select an
appropriate color-blind safe palette to use.

``` r
color.by = "Sex"
color.by = list('Body Site' = c("Saliva", "Stool"))
```



## Custom colors

You can map your metadata values to any color you'd like using the syntax 
in the following examples. You can use hexadecimal codes or any of the 657 
named R colors (type `colors()` for the list).

``` r
color.by = list('Sex' = c('Male' = "cyan", 'Female' = "#FFC83D"))
color.by = list('Age' = c("lightblue", "darkblue"))
color.by = list('Age' = list(colors = c("red", "blue"), range = c(30, NA)))
```


## Built-in palettes

A collection of color-blind friendly palettes are pre-defined in rbiom. They have been compiled from multiple sources:

* Crameri, Fabio (2023). *Scientific colour maps*. [website](https://www.fabiocrameri.ch/colourmaps/) &nbsp; [doi](https://doi.org/10.5281/zenodo.8409685)
* Frerebeau, Nicolas (2023). *khroma: Colour Schemes for Scientific Data Visualization*. [website](https://packages.tesselle.org/khroma/) &nbsp; [doi](https://doi.org/10.5281/zenodo.1472077)
* Neuwirth, Erich (2022). *ColorBrewer Palettes*. [website](http://colorbrewer2.org)
* Okabe, Masataka, and Ito, Key (2008). *Color Universal Design (CUD): How to Make Figures and Presentations That Are Friendly to Colorblind People*. [website](https://jfly.uni-koeln.de/color/)
* Tol, Paul (2021). *Colour Schemes*. [pdf](https://personal.sron.nl/~pault/data/colourschemes.pdf)

To use a specific one, assign its name to a metadata column.
``` r
# Most plots expect `color.by` to be a single categorical metadata field.
color.by = list('Body Site' = "bright")
color.by = list(
  'Body Site' = list('colors' = "bright", 'values' = c("Stool", "Saliva")) )

# Heatmap plots allow multiple metadata fields of any type.
color.by = list('Body Site' = "okabe", 'Age' = "bamako")
color.by = list(
  'Body Site' = list('colors' = "okabe", 'values' = c("Stool", "Saliva")),
  'Age'       = list('colors' = "bamako", 'range' = c(NA, 50)) )
```

### Discrete palettes

```{r cat_palettes, echo=FALSE, results='asis'}
cat(glue::glue_data(pals,
'<table>
<tr>{bright}{tableau}</tr>
<tr>{vibrant}{classic}</tr>
<tr>{okabe}{carto}</tr>
<tr>{r4}{tol}</tr>
<tr>{light}{tableau20}</tr>
<tr>{muted}{kelly}</tr>
<tr>{fishy}{alphabet}</tr>
<tr>{sub("<td>", "<td colspan=3>", polychrome)}</tr>
</table>'))
```


### Continuous palettes

```{r num_palettes, echo=FALSE, results='asis'}
cat(glue::glue_data(pals,
'<table>
<tr>{acton}{devon}{oslo}</tr>
<tr>{bam}{grays}{purples}</tr>
<tr>{bamako}{greens}{reds}</tr>
<tr>{batlow}{hawaii}{roma}</tr>
<tr>{berlin}{imola}{tofino}</tr>
<tr>{bilbao}{lajolla}{tokyo}</tr>
<tr>{broc}{lapaz}{turku}</tr>
<tr>{buda}{lisbon}{vanimo}</tr>
<tr>{cork}{nuuk}{vik}</tr>
<tr>{davos}{oranges}<td></td><td></td></tr>
</table>'))
```


### Reversing a palette

Prefix a palette name with a dash (`-`) to reverse the color order.
This is also helpful for assigning colors to a heatmap's main grid.
``` r
color.by = list('Age' = "-bamako")
grid = "-imola"
```






# Patterns

The patterned fills available in rbiom plots are made possible by the 
[ggpattern](https://coolbutuseless.github.io/package/ggpattern/) R package.

54 pattern names are defined by `gridpattern::names_magick`, however, the 
20 below are a good collection for most purposes. You can allow rbiom to 
choose a set of patterns, or you can explicitly define the mappings.


``` r
# One categorical metadata field may be specified.
pattern.by = "Body Site"
pattern.by = list('Body Site' = c("Stool", "Saliva"))
pattern.by = list('Body Site' = c('Stool' = "left45", 'Saliva' = "hs_cross"))

# Assign the same metadata field to color.by for colored patterns.
pattern.by = "Body Site", color.by = "Body Site"
```


```{r patterns, echo=FALSE}

patterns <- rbiom:::get_n_patterns(20)

ggplot2::ggplot(data.frame(pattern = patterns)) + 
  ggpattern::geom_tile_pattern(
    mapping      = ggplot2::aes(x = 1, y = 1, width = 1, height = 1, pattern_type = pattern), 
    stat         = "identity", 
    pattern      = "magick", 
    pattern_fill = "black", 
    fill         = "white", 
    color        = "black", 
    pattern_res  = 85 ) + 
  ggplot2::facet_wrap(~pattern) +
  ggpattern::scale_pattern_type_manual(
    guide  = "none",
    values = setNames(patterns, patterns) ) +
  ggplot2::theme_void() +
  ggplot2::theme(
    panel.spacing    = ggplot2::unit(10, "pt"),
    panel.background = ggplot2::element_rect(fill = "#222222", colour = NA),
    strip.background = ggplot2::element_rect(fill = "#222222", colour = NA),
    strip.text       = ggplot2::element_text(
      size   = 11, 
      color  = "white",
      face   = "bold",
      margin = ggplot2::margin(4,0,4,0, "pt") ))
```



# Shapes

Shapes can be given as per base R - numbers 0 through 17 for various shapes,
or the decimal value of an ascii character, e.g. 65:90 = A-Z; 97:122 = a-z to use 
letters instead of shapes on the plot. Text strings may used as well.


``` r
# One categorical metadata field may be specified.
shape.by = "Body Site"
shape.by = list('Body Site' = c("Stool", "Saliva"))
shape.by = list('Body Site' = c('Stool' = 7, 'Saliva' = 8))

# Assign the same metadata field to color.by for colored shapes.
shape.by = "Body Site", color.by = "Body Site"
```


```{r shapes, echo=FALSE, fig.asp = 0.25}
ggplot2::ggplot(data.frame(shape = c(0:17, 65, 90, 97, 122))) + 
  ggplot2::geom_point(
    mapping = ggplot2::aes(x = 1, y = 1, shape = shape),
    size    = 8,
    color   = "white" ) + 
  ggplot2::scale_shape_identity() +
  ggplot2::facet_wrap(~shape, nrow = 2) +
  ggplot2::theme_void() +
  ggplot2::theme(
    panel.spacing    = ggplot2::unit(10, "pt"),
    panel.background = ggplot2::element_rect(fill = "#222222", colour = NA),
    strip.background = ggplot2::element_rect(fill = "#222222", colour = NA),
    strip.text       = ggplot2::element_text(
      size   = 11, 
      color  = "white",
      face   = "bold",
      margin = ggplot2::margin(4,0,4,0, "pt") ))
```



# Facets

Faceting allows you to split your dataset into multiple sub-plots by one or 
more categorical metadata fields. For example:

```{r facets, fig.asp = 0.4, fig.align = 'center', out.width = "75%", fig.width = 6}
biom <- rarefy(hmp50)

adiv_boxplot(biom, color.by="Body Site", facet.by="Sex")
adiv_boxplot(
  biom     = biom, 
  color.by = "Sex", 
  facet.by = list('Body Site' = c("Saliva", "Stool")) )
```

## Auto-generated facets

Plotting functions often allow you to specify multiple ranks, taxa, metrics, 
etc which are automatically assigned their own facet dimension. You can still
add your own facets to these plots.

```{r auto_facets1, fig.asp = 0.3, fig.align = 'center', out.width = "75%", fig.width = 5}
bdiv_ord_plot(biom, layers = "p", bdiv = c("Bray-Curtis", "Euclidean"))
```

```{r auto_facets2, fig.asp = 0.55, fig.align = 'center', out.width = "75%", fig.width = 5}
bdiv_ord_plot(biom, layers = "p", bdiv = c("bray", "euc"), facet.by = "Sex")
```


# Limits

All the `*.by` parameters allow specifying a subset of samples to plot. 
`limit.by` does so without setting any other aesthetics. When multiple subsets 
are defined, the intersection is used for the plot.

``` r
# Only show samples from males whose age is >= 20 and <= 40.
limit.by = list('Sex' = "Male", 'Age' = c(20,40))

# Limit to these two body sites and age <= 35.
limit.by = list('Body Site' = c("Saliva", "Anterior nares"), 'Age' = c(NA,35))
```
