---
title: "Plot Types"
---

```{r, include = FALSE}
library(rbiom)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  R.options = list(
    pillar.print_min = 5,
    pillar.print_max = 5 ))
```


Plotting functions in rbiom fall into five categories:

| Category          | Functions                                             |
| ----------------- | ----------------------------------------------------- |
| Box Plots         | `adiv_boxplot()` `bdiv_boxplot()` `taxa_boxplot()`    |
| Correlation Plots | `adiv_corrplot()` `bdiv_corrplot()` `taxa_corrplot()` |
| Ordination Plots  | `bdiv_ord_plot()`                                     |
| Heatmaps          | `bdiv_heatmap()` `taxa_heatmap()` `plot_heatmap()`    |
| Stacked Bar Plots | `taxa_stacked()`                                      |


# Box Plots

Box plots are useful for visualizing a numeric outcome against one or more 
categorical predictors. The rbiom package provides dedicated functions for 
three numeric outcomes:

1. Alpha Diversity (shannon, simpson, etc) - `adiv_boxplot()`.
2. Beta Diversity (unifrac, bray-curtis, etc) - `bdiv_boxplot()`.
3. Taxa Abundance (phylum, genus, etc) - `taxa_boxplot()`.

You can then map categorical metadata variables to the x-axis, colors, 
patterns, shapes, and facets.
See [Mapping Metadata to Aesthetics](aes.html) for details.


## Layers

Despite being refered to as "box plots", rbiom's `*_boxplot()` functions 
support a wide range of graphical elements beyond box-and-whisker.

You can assign one or more of the following options to a box plot's `layers` 
parameter.

```{r, echo=FALSE}
biom <- rarefy(hmp50)
patchwork::wrap_plots(lapply(
  c("bar", "box", "violin", "dot", "strip", "crossbar", "errorbar", "linerange", "pointrange"), 
  FUN = function (layer) {
  adiv_boxplot(biom, layers = layer, color.by = "Sex", p.label = 0, y.limits = c(0, 4)) + 
    ggplot2::labs(title = layer, y = NULL, caption = NULL) +
    ggplot2::theme(legend.position = "none")
  }))
```

Unambiguous abbreviations are also accepted. For instance, 
`layers = c("box", "dot")` is equivalent to `layers = c("x", "d")` and 
`layers = "xd"`. Note that the single letter abbreviation for "box" is "x" 
("bar" is "b").


## Examples

```{r}
biom <- rarefy(hmp50, depth = 1000)

adiv <- adiv_boxplot(
  biom = biom, layers = "bes", x = "Sex", color.by = "Sex")
  
bdiv <- bdiv_boxplot(
  biom = biom, layers = "xd", x = "==Body Site", color.by = "Body Site", 
  pt.alpha = 0.2, pt.stroke = 0 )

taxa <- taxa_boxplot(
  biom = biom, layers = "p", x = "Body Site", p.label = 0, taxa = 3, 
  facet.ncol = 1, facet.strip.position = "right" )

patchwork::wrap_plots(
  lapply(list(adiv, bdiv, taxa), function (p) {
    p + 
      ggplot2::labs(x = NULL, caption = NULL) + 
      ggplot2::theme(legend.position = "none") }))
```


## Statistics

The categorical groups defined by `x`, `color.by`, etc are used to calculate 
non-parametric statistics with the Mann-Whitney or Kruskal-Wallis algorithms. 
See [Statistics](statistics.html) for details.




# Correlation Plots

Correlation plots are designed to visualize the relationship between a numeric 
outcome and a numeric predictor, optionally with additional categorical 
predictors. The rbiom package provides dedicated functions for 
three numeric outcomes:

1. Alpha Diversity (shannon, simpson, etc) - `adiv_corrplot()`.
2. Beta Diversity (unifrac, bray-curtis, etc) - `bdiv_corrplot()`.
3. Taxa Abundance (phylum, genus, etc) - `taxa_corrplot()`.

You can then map categorical metadata variables to the x-axis, colors, and 
facets. See [Mapping Metadata to Aesthetics](aes.html) for details.



## Layers

You can assign one or more of the following options to a correlation plot's 
`layers` parameter.

```{r, echo=FALSE}
biom <- rarefy(hmp50)
patchwork::wrap_plots(lapply(
  c("trend", "confidence", "point", "residual", "name"), 
  FUN = function (layer) {
  adiv_corrplot(biom, x = "Age", layers = layer, color.by = "Sex", test = "none", y.limits = c(0, 50)) + 
    ggplot2::labs(title = layer, y = NULL, caption = NULL) +
    ggplot2::theme(legend.position = "none")
  }))
```

Unambiguous abbreviations are also accepted. For instance, 
`layers = c("trend", "point")` is equivalent to `layers = c("t", "p")` and 
`layers = "tp"`.



## Examples

```{r}
biom <- rarefy(hmp50, depth = 1000)

adiv <- adiv_corrplot(biom, layers = "ts", x = "BMI")
  
bdiv <- bdiv_corrplot(
  biom = biom, layers = "tc", x = "BMI", color.by = "==Body Site", 
  limit.by = list("Body Site" = c("Buccal mucosa", "Stool")) )

taxa <- taxa_corrplot(
  biom = biom, layers = "tsr", x = "Age", taxa = 3, color.by = "Body Site", 
  limit.by = list("Body Site" = c("Mid vagina", "Saliva", "Stool")),
  facet.ncol = 1, facet.strip.position = "right" )

patchwork::wrap_plots(
  guides = "collect",
  lapply(list(adiv, bdiv, taxa), function (p) {
    p + ggplot2::labs(caption = NULL) }))
```




# Ordination Plots



## Layers

You can assign one or more of the following options to a ordination plot's 
`layers` parameter.

```{r, echo=FALSE}
biom <- hmp50 %>% subset(`Body Site` %in% c("Saliva", "Buccal mucosa")) %>% rarefy()
patchwork::wrap_plots(lapply(
  c("point", "spider", "ellipse", "name", "mean", "taxon", "arrow"), 
  FUN = function (layer) {
  bdiv_ord_plot(biom, layers = layer, color.by = "Body Site", test = "none") + 
    ggplot2::ggtitle(layer) +
    ggplot2::theme(legend.position = "none")
  }))
```

Unambiguous abbreviations are also accepted. For instance, 
`layers = c("point", "ellipse")` is equivalent to `layers = c("p", "e")` and 
`layers = "pe"`.

The layers `c("point", "spider", "ellipse", "name")` apply to samples.
The layers `c("mean", "taxon", "arrow")` apply to the taxa.



## Examples

```{r}
biom <- rarefy(hmp50)

p1 <- bdiv_ord_plot(biom, layers = "pse", color.by = "Body Site") +
        ggplot2::theme(legend.position = "none")

p2 <- bdiv_ord_plot(biom, layers = "emt", color.by = "Body Site")

patchwork::wrap_plots(p1, p2, guides = "collect")
```




# Heatmaps

Visualizing a large grid of values is a job for heatmaps. The generic 
`plot_heatmap()` function accepts any matrix, while the two common use cases 
below operate on a biom object.

1. Beta Diversity (unifrac, bray-curtis, etc) - `bdiv_heatmap()`.
2. Taxa Abundance (phylum, genus, etc) - `taxa_heatmap()`.



## Examples

```{r}
biom <- rarefy(hmp50)

bdiv <- bdiv_heatmap(biom, color.by = c("Age", "Body Site"), asp = 0.4)
taxa <- taxa_heatmap(biom, color.by = c("Age", "Body Site"), asp = 0.4)

patchwork::wrap_plots(bdiv, taxa, ncol = 1, guides = "collect")
```




# Stacked Bar Plots

Taxa stacked bar plots show the same information as a taxa heatmap. Both have 
their own advantages.


## Examples

```{r}
taxa_stacked(rarefy(hmp50), taxa = 10)
```
