---
title: "Statistics"
---

```{r, include = FALSE}
library(rbiom)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  R.options = list(
    pillar.print_min = 5,
    pillar.print_max = 5 ))
```


# Quick Start

The rbiom package includes many statistical functions. If you have an rbiom object, you can use dedicated functions for alpha diversity, beta diversity, and taxa abundance. Otherwise, look to the generic functions that operate on any data.frame or distance matrix.





## rbiom Object Functions

Your metadata field and microbiome property of interest will determine which 
rbiom function to use.

<table>
<tr>
<th rowspan=2 class="border-0 border-end border-bottom align-middle">Metadata<br>Field</th>
<th colspan=3 class="border-0" align="center">Microbiome Property</th>
</tr><tr>
<th class="border-bottom">Alpha Diversity<br><span class="eg">Shannon, Simpson</span></th>
<th class="border-bottom">Beta Diversity<br><span class="eg">UniFrac, Jaccard</span></th>
<th class="border-bottom">Taxa Abundance<br><span class="eg">Phylum, Genus</span></th>
</tr><tr>
<th class="border-end">Categorical<br><span class="eg">Sex, Body Site</span></th>
<td>`adiv_boxplot()`</td>
<td>`bdiv_boxplot()`<br>`bdiv_ord_plot()`</td>
<td>`taxa_boxplot()`</td>
</tr><tr>
<th class="border-end">Numeric<br><span class="eg">Age, BMI</span></th>
<td>`adiv_corrplot()`</td>
<td>`bdiv_corrplot()`</td>
<td>`taxa_corrplot()`</td>
</tr><tr>
<th class="border-end"><i>Any</i></th>
<td>`adiv_stats()`</td>
<td>`bdiv_stats()`</td>
<td>`taxa_stats()`</td>
</tr>
</table>

For instance, to explore the effect of Body Site (a categorical metadata field) 
on Shannon Diversity (an alpha diversity metric), we'd use `adiv_boxplot()` to 
produce a plot with statistics, or `adiv_stats()` if we only want the stats.





## Generic Functions

If your data is in a data.frame (or tibble), use:

```{r, echo=FALSE, results='asis'}
cat(glue::glue('
  <table class="table-borderless">
  <tr>{stats_boxplot}{stats_corrplot}{stats_table}</tr>
  </table>',
  .transformer = function (text, ...) {
    sprintf("<td><code><a href='../reference/%s.html'>%s()</a></code></td>", text, text)
  }))
```

Or, for a distance matrix:

```{r, echo=FALSE, results='asis'}
cat(glue::glue('
  <table class="table-borderless">
  <tr>{distmat_stats}</tr>
  </table>',
  .transformer = function (text, ...) {
    sprintf("<td><code><a href='../reference/%s.html'>%s()</a></code></td>", text, text)
  }))
```





# Statistics Table

The function `stats_table()` and functions ending in "_stats" (e.g. `adiv_stats()`, `distmat_stats()`) will return a statistics table. Functions with "plot" in their name (e.g. `adiv_boxplot()`, `stats_corrplot()`) will return a plot object `p`. You can access the plot's associated statistics table with `p$stats`. 


```{r}
p <- adiv_boxplot(
  biom     = rarefy(hmp50),           # Dataset as an rbiom object
  adiv     = c("Shannon", "Simpson"), # Alpha diversity metrics
  stat.by  = "Body Site",             # Statistical groups
  facet.by = "Sex" )                  # Split data prior to stats

p$stats
```

This table has the following information:





## Model

In the upper-left corner, we see ``Model:    kruskal.test(.diversity ~ `Body Site`)``. This tells us that the underlying test used here is base R's `kruskal.test()`, grouping values of `.diversity` by `` `Body Site` ``. In rbiom, `.diversity` is the default name for the column containing alpha diversity values.





## Multiple Comparisons

The above table has four rows, meaning we ran data through the model four times. In this case, it was two alpha diversity metrics times two `Sex` facets. The first two columns of the table (`Sex` and `.adiv`) tell us which row is for each combination. Additionally, the `.adj.p` column is the `.p.val` adjusted for multiple comparisons.





## Hypothesis

The three columns `.stat`, `.h1`, and `.p.val` together show the hypothesis and its outcome. Above, we're asking if `.stat` is greater than zero. When `.p.val` is less than 0.05 we can say that it is.

That's the very simplified explanation. 

Recall from your statistic classes that you are testing a null hypothesis H<sub>0</sub> against an alternate hypothesis H<sub>1</sub>. In this case, our null hypothesis is that the Kruskal-Wallis statistic is zero (`.stat == 0`), indicating all `` `Body Site` `` groups have similar alpha diversity. The p-value is the probability that the null hypothesis is correct (a p-value of 0.6 is interpreted as a 60% chance that `.stat == 0`). When the p-value is below a certain value (usually 0.05) we accept the alternative hypothesis instead - in this case that `.stat > 0`, meaning `.diversity` does vary by `` `Body Site` ``.





## Terms of Interest

The last two columns are:

* `.n` - The number of samples. After rarefying, we have 30 females and 19 males.
* `.df` - Degrees of freedom. One less than the number of groups. The HMP50 dataset has five body sites: Anterior nares, buccal mucosa, mid vagina, saliva, and stool. However, the male facets don't have any mid vagina samples.




## Data and Code

You can inspect the data passed to `kruskal.test()` by looking at `p$data`. Additionally, the code used to generate the statistics is available in `p$stats$code`. These attributes can help you reproduce and customize your data analysis.

```{r}
p$data

p$stats$code
```

When generating statistics with `st <- adiv_stats()`, `bdiv_stats()`, or `taxa_stats()`, the data and code are in `st$data` and `st$code`, respectively. For `st <- stats_table(df = df)`, the data and code are in `df` and `st$code`, respectively.



## Column Reference

Care has been taken to keep rbiom's statistics tables consistent across functions. However, some tables will provide more information when it is available from the underlying statistical function.

Below is a quick reference guide to all columns that may appear in an rbiom statistics table.

| **Field**      | **Description**                                       |
| -------------- | ----------------------------------------------------- |
| `.stat`        | Wilcoxon or Kruskal-Wallis rank sum statistic.        |
| `.mean`        | Estimated marginal mean. See `emmeans::emmeans()`.    |
| `.mean.diff`   | Difference in means.                                  |
| `.slope`       | Trendline slope. See `emmeans::emtrends()`.           |
| `.slope.diff`  | Difference in slopes.                                 |
| `.h1`          | Alternate hypothesis.                                 |
| `.p.val`       | Probability that null hypothesis is correct.          |
| `.adj.p`       | `.p.val` after adjusting for multiple comparisons.    |
| `.effect.size` | Effect size. See `emmeans::eff_size()`.               |
| `.lower`       | Confidence interval lower bound.                      |
| `.upper`       | Confidence interval upper bound.                      |
| `.se`          | Standard error.                                       |
| `.n`           | Number of samples.                                    |
| `.df`          | Degrees of freedom.                                   |
| `.t.ratio`     | *(.mean, .mean.diff, .slope, or .slope.diff)* / `.se` |
| `.z`           | Std. effect size. See `vegan::summary.permustats()`.  |
| `.r.sqr`       | Percent of variation explained by the model.          |
| `.adj.r`       | `.r.sqr`, taking degrees of freedom into account.     |
| `.aic`         | Akaike Information Criterion (predictive models).     |
| `.bic`         | Bayesian Information Criterion (descriptive models).  |
| `.loglik`      | Log-likelihood goodness-of-fit score.                 |
| `.fit.p`       | P-value for observing this fit by chance.             |

The `.h1` field will always come immediately after the column it is testing against.




# Plot Output

Visualizations are one of the best ways to identify correlations in your 
dataset. If you can see a trend with your eyes, then you're on the right track. 
The statistics-supported plotting functions in rbiom are ordination plots, box 
plots, and correlation plots.





## Ordination Plots

Statistics for ordination plots are the most straight-forward. Set a 
categorical metadata field to the `stat.by` parameter to test whether 
inter-sample distances are correlated with that variable.

```{r}
p <- bdiv_ord_plot(
  biom    = rarefy(hmp50), 
  stat.by = "Body Site", 
  bdiv    = c("Jaccard", "Bray-Curtis"), 
  ord     = c("PCoA", "UMAP") )
p
p$stats

p$stats$code
```

The plot subtitles have the summary statistics. Additionally, `p$stats` 
contains a tibble data.frame with the full statistics table, and `p$stats$code` 
shows the R commands for reproducing the statistics outside of rbiom.

Note that the ordination statistics are not dependent on the ordination, only 
the distance metric. This is because the statistics are based on beta diversity 
distances which are computed prior to ordination.

By default, `bdiv_ord_plot()` applies the perMANOVA test. You can change this 
to MRPP by specifying `test="mrpp"`. Details on the available tests are below.

| Test      | Function           | Method                                                      |
| --------- | ------------------ | ----------------------------------------------------------- |
| `adonis2` | `vegan::adonis2()` | Permutational Multivariate Analysis of Variance (perMANOVA) |
| `mrpp`    | `vegan::mrpp()`    | Multiple Response Permutation Procedure (MRPP)              |





## Box Plots

Statistics on box plots will automatically toggle between pairwise and 
group-wise statistics based on the values of `x` and `stat.by`: `x` controls 
pairwise and `stat.by` controls group-wise. You can set `x` and `stat.by` to 
the same categorical metadata field to get colored pairwise statistics, or set 
them to different categorical metadata fields to get multiple group-wise 
statistics per plot.

```{r}
biom <- rarefy(hmp50) %>% 
  subset(`Body Site` %in% c('Saliva', 'Stool', 'Buccal mucosa'))

p1 <- adiv_boxplot(biom, x = "Body Site", stat.by = NULL)
p2 <- adiv_boxplot(biom, x = NULL,        stat.by = "Body Site")
p3 <- adiv_boxplot(biom, x = "Body Site", stat.by = "Body Site")
p4 <- adiv_boxplot(biom, x = "Sex",       stat.by = "Body Site")

plots <- list(
  p1 + ggplot2::labs(subtitle = 'x = "Body Site", stat.by = NULL'), 
  p2 + ggplot2::labs(subtitle = 'x = NULL, stat.by = "Body Site"'), 
  p3 + ggplot2::labs(subtitle = 'x = "Body Site", stat.by = "Body Site"'), 
  p4 + ggplot2::labs(subtitle = 'x = "Sex", stat.by = "Body Site"') ) %>%
  lapply(`+`, ggplot2::labs(x = NULL, y = NULL, caption = NULL)) %>%
  lapply(`+`, ggplot2::theme(plot.subtitle = ggplot2::element_text(size = 10)))

patchwork::wrap_plots(plots, guides = "collect")
```

Above, the lower-left plot is annotated with pairwise statistics while the two 
on the right have group-wise statistics. As with other plots, you can find the 
full statistics tables and reproducible R code in the plot attributes.

```{r}
p3$stats

p2$stats

p2$stats$code
```

Internally, rbiom uses the non-parametric functions listed below.

| Test          | Function                | Method                                                   |
| ------------- | ----------------------- | -------------------------------------------------------- |
| pairwise      | `stats::wilcox.test()`  | Two-sample Wilcoxon Rank Sum Test, aka Mann-Whitney Test |
| group-wise    | `stats::kruskal.test()` | Kruskal-Wallis Rank Sum Test                             |






## Correlation Plots

For an in-depth description of correlation plots, see the 
[rbiom regression article](regression.html).

Depending on the arguments given to `stat.by` and `test`, you can test:

* Is the mean non-zero?
* Do means vary by group?
* Is the slope (trendline) non-horizontal?
* Do slopes vary by group?


```{r}
biom <- gems %>%
  subset(diarrhea == "Control") %>%
  subset(country %in% c("Bangladesh", "Kenya")) %>%
  rarefy()

p1 <- adiv_corrplot(biom, x = "age", test = "emmeans")
p2 <- adiv_corrplot(biom, x = "age", test = "emmeans", stat.by = "country")
p3 <- adiv_corrplot(biom, x = "age", test = "emtrends")
p4 <- adiv_corrplot(biom, x = "age", test = "emtrends", stat.by = "country")

plots <- list(
  p1 + ggplot2::labs(subtitle = 'Is the mean non-zero?'), 
  p2 + ggplot2::labs(subtitle = 'Do means vary by group?'), 
  p3 + ggplot2::labs(subtitle = 'Is the slope non-horizontal?'), 
  p4 + ggplot2::labs(subtitle = 'Do slopes vary by group?')) %>%
  lapply(`+`, ggplot2::labs(x = NULL, y = NULL, caption = NULL))

patchwork::wrap_plots(plots, guides = "collect")
```




# Background


## Normality

A normal distribution is visualized as a "bell curve", where values further 
from the mean are observed less often. Microbial abundances do not follow this 
pattern; it's common to observe high or low abundances more often than a 
"medium" abundance.

```{r, fig.asp = 0.4}
library(ggplot2)

patchwork::wrap_plots(
  widths = c(1, 1.5),

  ggplot() + 
    geom_histogram(aes(x=rnorm(1000)), bins = 10) + 
    ggtitle("Normal Distribution"),
    
  ggplot(data = taxa_table(rarefy(hmp50), taxa = 4)) + 
    geom_histogram(aes(x=.abundance), bins = 10) + 
    facet_wrap(".taxa") + 
    ggtitle("Genera Abundance Distributions")
)
```

To compensate for this non-normality, rbiom uses the following non-parametric 
tests for categorical variables that are based on ranking or permutations.

| Test                    | Function                | Used For                   |
| ----------------------- | ----------------------- | -------------------------- |
| Wilcoxon Rank-Sum       | `stats::wilcox.test()`  | Pairwise boxplot           |
| Kruskal-Wallis Rank Sum | `stats::kruskal.test()` | Groupwise boxplot          |
| Permutational MANOVA    | `vegan::adonis2()`      | `bdiv_ord_plot()` clusters |

For correlation/regression analysis, rbiom provides diagnostic plots to determine when residual distributions are cause for concern. To enable this feature, set `check = TRUE`.

```{r}
adiv_corrplot(biom, x = "age", test = "emmeans", check = TRUE)
```



Further reading:

* [Applied Multivariate Statistics in R](https://uw.pressbooks.pub/appliedmultivariatestatistics/): 
[PERMANOVA](https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/permanova/), 
[Comparison of Techniques](https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/comparison-of-techniques/)
* [R Applications — Part 1: Simple Linear Regression](https://medium.com/datasciencearth/r-applications-part-1-simple-linear-regression-ef5a0e19a05d), 
particularly the section on [assumption control](https://medium.com/datasciencearth/r-applications-part-1-simple-linear-regression-ef5a0e19a05d#a6f6).




## Compositionality

Compositional data arises when the counts don't represent the entire 
population. In microbiome studies, the number of microbes that get sequenced 
is far less than the number of microbes from where the sample was collected. 
Articles by [Gloor et al](https://doi.org/10.3389/fmicb.2017.02224) and 
[McMurdie and Holmes](https://doi.org/10.1371/journal.pcbi.1003531) 
propose the use of their analysis tools (
[ALDEx2](https://doi.org/doi:10.18129/B9.bioc.ALDEx2) and
[metagenomeSeq](https://doi.org/doi:10.18129/B9.bioc.metagenomeSeq), 
respectively) to apply the proper statistical methods for this situation. 
Conversely, rbiom does not correct for compositionality. This is because 
correcting for compositionality introduces extra noise into the dataset and 
severely limits the selection of metrics and visualizations, typically 
without any significant benefit to analysis.

