---
title: "Statistics"
---

```{r, include = FALSE}
library(rbiom)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  R.options = list(
    pillar.print_min = 5,
    pillar.print_max = 5 ))
```


# Introduction

Visualizations are one of the best ways to identify correlations in your 
dataset. If you can see a trend with your eyes, then you're on the right track. 
Rbiom's plotting functions are 

```{r, echo=FALSE, results='asis'}
cat(glue::glue('
  <table class="table-borderless">
  <tr>{adiv_boxplot}{bdiv_corrplot}{rare_barplot}{taxa_boxplot}</tr>
  <tr>{adiv_corrplot}{bdiv_heatmap}{rare_corrplot}{taxa_corrplot}</tr>
  <tr>{bdiv_boxplot}{bdiv_ord_plot}{taxa_barplot}{taxa_heatmap}</tr>
  </table>',
  .transformer = function (text, ...) {
    sprintf("<td><code><a href='../reference/%s.html'>%s()</a></code></td>", text, text)
  }))
```

The `*_boxplot()`, `*_corrplot()`, and `bdiv_ord_plot()` functions will 
automatically add p-values to your figures whenever possible. The ggplot object 
they return has `$data`, `$code`, `$stats`, and `$stats$code` attributes you 
can use to automate, reproduce, and customize your figures.

To generate statistics without creating a plot, use one of the following:

```{r, echo=FALSE, results='asis'}
cat(glue::glue('
  <table class="table-borderless">
  <tr>{adiv_stats}{bdiv_stats}{distmat_stats}{taxa_stats}</tr>
  </table>',
  .transformer = function (text, ...) {
    sprintf("<td><code><a href='../reference/%s.html'>%s()</a></code></td>", text, text)
  }))
```



\cr

# Quick Start

Your metadata field of interest and microbiome property will determine which 
rbiom function to use.

<table>
<tr>
<th rowspan=2 class="border-0 border-end border-bottom align-middle">Metadata<br>Property</th>
<th colspan=3 class="border-0" align="center">Microbiome Property</th>
</tr><tr>
<th class="border-bottom">Alpha Diversity<br><span class="eg">Shannon, Simpson</span></th>
<th class="border-bottom">Beta Diversity<br><span class="eg">UniFrac, Jaccard</span></th>
<th class="border-bottom">Taxa Abundance<br><span class="eg">Phylum, Genus</span></th>
</tr><tr>
<th class="border-end">Categorical<br><span class="eg">Sex, Body Site</span></th>
<td>`adiv_boxplot()`</td>
<td>`bdiv_boxplot()`<br>`bdiv_ord_plot()`</td>
<td>`taxa_boxplot()`</td>
</tr><tr>
<th class="border-end">Numeric<br><span class="eg">Age, BMI</span></th>
<td>`adiv_corrplot()`</td>
<td>`bdiv_corrplot()`</td>
<td>`taxa_corrplot()`</td>
</tr><tr>
<th class="border-end"><i>Any</i></th>
<td>`adiv_stats()`</td>
<td>`bdiv_stats()`</td>
<td>`taxa_stats()`</td>
</tr>
</table>

For instance, to explore the effect of Body Site (a categorical metadata field) 
on Shannon Diversity (an alpha diversity metric), we'd use `adiv_boxplot()` to 
produce a plot with statistics, or `adiv_stats()` if we only want the stats.

Although all the functions have important differences, the statistical methods 
they employ can be grouped by the three plot types: ordination plot, box plot, 
and correlation plot.


\cr

# Ordination Plots

Statistics for ordination plots are the most straight-forward. Set a 
categorical metadata field to the `color.by` parameter to test whether 
inter-sample distances are correlated with that variable.

```{r}
p <- bdiv_ord_plot(
  biom     = rarefy(hmp50), 
  color.by = "Body Site", 
  bdiv     = c("Jaccard", "Bray-Curtis"), 
  ord      = c("PCoA", "UMAP") )
p
p$stats

p$stats$code
```

The plot subtitles have the summary statistics. Additionally, `p$stats` 
contains a tibble data.frame with the full statistics table, and `p$stats$code` 
shows the R commands for reproducing the statistics outside of rbiom.

Note that the ordination statistics are not dependent on the ordination, only 
the distance metric. This is because the statistics are based on beta diversity 
distances which are computed prior to ordination.

By default, `bdiv_ord_plot()` applies the perMANOVA test. You can change this 
to MRPP by specifying `test="mrpp"`. Details on the available tests are below.

| Test      | Function           | Method                                                      |
| --------- | ------------------ | ----------------------------------------------------------- |
| `adonis2` | `vegan::adonis2()` | Permutational Multivariate Analysis of Variance (perMANOVA) |
| `mrpp`    | `vegan::mrpp()`    | Multiple Response Permutation Procedure (MRPP)              |



\cr

# Box Plots

Statistics on box plots will automatically toggle between pairwise and 
group-wise statistics based on the values of `x` and `color.by`: `x` controls 
pairwise and `color.by` controls group-wise. You can set `x` and `color.by` to 
the same categorical metadata field to get colored pairwise statistics, or set 
them to different categorical metadata fields to get multiple group-wise 
statistics per plot.

```{r}
biom <- rarefy(hmp50) %>% 
  subset(`Body Site` %in% c('Saliva', 'Stool', 'Buccal mucosa'))

p1 <- adiv_boxplot(biom, x = "Body Site")
p2 <- adiv_boxplot(biom, color.by = "Body Site")
p3 <- adiv_boxplot(biom, x = "Body Site", color.by = "Body Site")
p4 <- adiv_boxplot(biom, x = "Sex", color.by = "Body Site")

plots <- list(
  p1 + ggplot2::labs(subtitle = 'x = "Body Site"'), 
  p2 + ggplot2::labs(subtitle = 'color.by = "Body Site"'), 
  p3 + ggplot2::labs(subtitle = 'x = color.by = "Body Site"'), 
  p4 + ggplot2::labs(subtitle = 'x = "Sex", color.by = "Body Site"')) %>%
  lapply(`+`, ggplot2::labs(x = NULL, y = NULL, caption = NULL))

patchwork::wrap_plots(plots, guides = "collect")
```

Above, the two plots on the left are annotated with pairwise statistics while 
the two on the right have group-wise statistics. As with other plots, you can 
find the full statistics tables and reproducible R code in the plot attributes.

```{r}
p1$stats

p2$stats

p2$stats$code
```

Internally, rbiom uses the non-parametric functions listed below.

| Test          | Function                | Method                                                   |
| ------------- | ----------------------- | -------------------------------------------------------- |
| pairwise      | `stats::wilcox.test()`  | Two-sample Wilcoxon Rank Sum Test, aka Mann-Whitney Test |
| group-wise    | `stats::kruskal.test()` | Kruskal-Wallis Rank Sum Test                             |




\cr

# Background


## Normality

A normal distribution is visualized as a "bell curve", where values further 
from the mean are observed less often. Microbial abundances do not follow this 
pattern; it's common to observe high or low abundances more often than a 
"medium" abundance.

```{r}
library(ggplot2)

patchwork::wrap_plots(
  widths = c(1, 1.5),

  ggplot() + 
    geom_histogram(aes(x=rnorm(1000)), bins = 10) + 
    ggtitle("Normal Distribution"),
    
  ggplot(data = taxa_table(rarefy(hmp50), taxa = 4)) + 
    geom_histogram(aes(x=.abundance), bins = 10) + 
    facet_wrap(".taxa") + 
    ggtitle("Genera Abundance Distributions")
)
```

To compensate for this non-normality, rbiom uses the following non-parametric 
tests for categorical variables that are based on ranking or permutations.

| Test                    | Function                | Used For                   |
| ----------------------- | ----------------------- | -------------------------- |
| Wilcoxon Rank-Sum       | `stats::wilcox.test()`  | Pairwise boxplot           |
| Kruskal-Wallis Rank Sum | `stats::kruskal.test()` | Groupwise boxplot          |
| Permutational MANOVA    | `vegan::adonis2()`      | `bdiv_ord_plot()` clusters |

For correlation/regression analysis, rbiom uses estimated marginal means 
(`emmeans::emmeans()`) on rank-transformed values.

Further reading:

* [Applied Multivariate Statistics in R](https://uw.pressbooks.pub/appliedmultivariatestatistics/): 
[PERMANOVA](https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/permanova/), 
[Comparison of Techniques](https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/comparison-of-techniques/)
* [R Applications — Part 1: Simple Linear Regression](https://medium.com/datasciencearth/r-applications-part-1-simple-linear-regression-ef5a0e19a05d), 
particularly the section on [assumption control](https://medium.com/datasciencearth/r-applications-part-1-simple-linear-regression-ef5a0e19a05d#a6f6).

\cr


## Compositionality

Compositional data arises when the counts don't represent the entire 
population. In microbiome studies, the number of microbes that get sequenced 
is far less than the number of microbes from where the sample was collected. 
Articles by [Gloor et al](https://doi.org/10.3389/fmicb.2017.02224) and 
[McMurdie and Holmes](https://doi.org/10.1371/journal.pcbi.1003531) 
propose the use of their analysis tools (
[ALDEx2](https://doi.org/doi:10.18129/B9.bioc.ALDEx2) and
[metagenomeSeq](https://doi.org/doi:10.18129/B9.bioc.metagenomeSeq), 
respectively) to apply the proper statistical methods for this situation. 
Conversely, rbiom does not correct for compositionality. This is because 
correcting for compositionality introduces extra noise into your dataset and 
severely limits your selection of metrics and visualizations, typically 
without any significant benefit to analysis.

